What We Eval in the Shadows

### 2021-08-12 paper/main.tex
Fix typo
PR section 3

In this work, we significantly expand that study. First, our corpus covers
\CranPackages packages (about the whole of CRAN at time of writing) with
\PkgHitEvalCallSites \eval call sites and \packageAllcallsRnd \eval calls.
Next, we track the number of information such as the shapes of expressions
passed to \eval, their provenance, complexity, and side effects. This enables
us to have an accurate and detailed picture of the use of \eval in the R
ecosystem. We use this to provide both quantitative and qualitative views on
\eval, including several use cases to demonstrate the popularity, diversity and
powerfulness of \eval in R.

@@

\begin{compactenum}[(1)]
\item \emph{Download.} Packages are downloaded from CRAN, for Kaggle a web
  crawler retrieves code, and a command-line tool gets data. Installation is
  complicated by native dependencies which sometimes have to be resolved
  manually.
\item \emph{Extract.} The \genthat tool extracts all runnable code snippets and
  turns each of these into a self-standing program~\cite{issta18}, \k{knitr}
  extracts code from notebooks. Each extracted script is instrumented with calls
  to our dynamic analyzer in a way that avoid recording calls to \eval
  originating from our execution harness.
\item \emph{Trace.} Script run with a modified R interpreter to capture
  execution traces. Each script is run in its own process with the GNU R
  compiler turned off to avoid recording its execution. CRAN scripts run twice,
  once to capture \eval calls originating from package code and a second time to
  capture calls coming from base libraries.
\item \emph{Analyze.} Analysis outputs are merged, cleaned, and summarized.
  RMarkdown notebooks process the summarized data to generate all graphs (PDF
  from \k{ggplot2}) and numbers (exported as \LaTeX macros) appearing in the
  paper.
\end{compactenum}

 The pipeline runs in parallel~\cite{GNUparallel} orchestrated by a
Makefile. Servers have identical environments thanks to a docker image with all
dependencies installed.

@@

The tracer that performs dynamic analysis of R scripts is built on top of
\rdyntrace, an extended R 4.0.2 virtual machine that exposes callbacks for
various runtime events~\cite{oopsla19b}. The tracer registers callbacks to all
variants of \eval and a few additional functions to locate the origin of \eval
arguments (it taints the results of calls to \k{parse} and \k{match.call}). It
captures dynamic code loading, as well as variable definition and assignment,
allowing us to record side effects that happen in environments while evaluating
code in \eval. A challenge was that lack of source code references for
expression that are not within block surrounded by braces. This is unfortunately
not easily fixed. We extend our dynamic analysis tool to attach synthetic source
code references to all \eval call sites, but the approach fails in some edge
cases. For performance reasons, the tracer is an R package written in C++ (3.2K
LOC) and R (1.3K LOC). While in theory the implementation should be
straightforward, not so in practice. Lazy evaluation requires delaying the
processing of arguments until (and if) they are forced. Accounting for
side-effects performed in an \eval is complicated by the fact that R is
implemented in a mixture of R and C, and that the language implementation can
call \eval. Lastly, large code bases exercise many edge cases of the highly
underspecified R behavior.

\paragraph{Limitations.} Even with the extension described above, there are
\PkgUndefinedRnd \eval calls without source information (\PkgUndefinedRatio of
all \eval calls). This happen when \eval is an argument to a higher-order
function or when it is called from native code. An other limitation is that we
do not record calls to the native \eval. Neither of these limitations should
invalidate our conclusions.

>>>>

In this work, we significantly expand that study. First, our corpus covers
\CranPackages packages (about the whole of CRAN at time of writing) with
\PkgHitEvalCallSites \eval call sites and \packageAllcallsRnd \eval calls.
Next, we track a number of information such as the shapes of expressions
passed to \eval, their provenance, complexity, and side effects. This enables
us to have an accurate and detailed picture of the use of \eval in the R
ecosystem. We use this to provide both quantitative and qualitative views on
\eval, including several use cases to demonstrate the popularity, diversity and
powerfulness of \eval in R.

@@

\begin{compactenum}[(1)]
\item \emph{Download.} Packages are downloaded from CRAN, for Kaggle a web
  crawler retrieves code, and a command-line tool gets data. Installation is
  complicated by native dependencies which sometimes have to be resolved
  manually.
\item \emph{Extract.} The \genthat tool extracts all runnable code snippets and
  turns each of these into a self-standing program~\cite{issta18}, \k{knitr}
  extracts code from notebooks. Each extracted script is instrumented with calls
  to our dynamic analyzer in a way that avoid recording calls to \eval
  originating from our execution harness.
\item \emph{Trace.} Script run with a modified R interpreter to capture calls
  to \eval in CRAN packages. Each script is run in its own process with the GNU
  R compiler turned off to avoid recording its execution.
\item \emph{Analyze.} Analysis outputs are merged, cleaned, and summarized.
  RMarkdown notebooks process the summarized data to generate all graphs (PDF
  from \k{ggplot2}) and numbers (exported as \LaTeX macros) appearing in the
  paper.
\end{compactenum}

\medskip The pipeline runs in parallel~\cite{GNUparallel} orchestrated by a
Makefile. Servers have identical environments thanks to a docker image with all
dependencies installed.

@@

The tracer that performs dynamic analysis of R scripts is built on top of
\rdyntrace, an extended R 4.0.2 virtual machine that exposes callbacks for
various runtime events~\cite{oopsla19b}. The tracer registers callbacks to all
variants of \eval and a few additional functions to locate the origin of \eval
arguments. It captures dynamic code loading, as well as variable definition and
assignment, allowing us to record side effects that happen in environments
while evaluating code in \eval. A challenge was that lack of source code
references for expression that are not within block surrounded by braces. This
is unfortunately not easily fixed. We extend our dynamic analysis tool to
attach synthetic source code references to all \eval call sites, but the
approach fails in some edge cases. For performance reasons, the tracer is an R
package written in C++ (3.2K LOC) and R (1.3K LOC). While in theory the
implementation should be straightforward, not so in practice. Lazy evaluation
requires delaying the processing of arguments until (and if) they are forced.
Accounting for side-effects performed in an \eval is complicated by the fact
that R is implemented in a mixture of R and C, and that the language
implementation can (and does) call \eval. Lastly, large code bases exercise
many edge cases of the highly underspecified R behavior.

\paragraph{Limitations.} Even with the extension described above, there are
\PkgUndefinedRnd \eval calls without source information (\PkgUndefinedRatio of
all \eval calls). This happen when \eval is an argument to a higher-order
function or when it is called from native code. Another limitation is that we
do not record calls to the native \eval. Neither of these limitations should
invalidate our conclusions.
